Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-10-09 analysis/paper/paper.Rmd
Ongoing - manuscript and method tweaking.

resultScenario = "S10"
resultSet ="cmult6ay5aSf1"
figSet = c("diffs","distScns","EVSI","power","EVPIall")

@@

```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios. Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%). These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, and not intended to reflect real landscape change. In all scenarios, the projection period begins in 2024, and the monitoring period occurs prior to the projection period.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay5aSf1.png"))

@@

To investigate the effectiveness of various monitoring strategies for
reducing uncertainty about population viability, and to show how this
varies with landscape condition, we begin with a model for simulating
example population dynamics and observations of those dynamics. These
example population trajectories are informed by prior information about
the distribution of possible parameter values and outcomes from the
national analysis of demographic-disturbance relationships
[@johnson_science_2020] so we call it the "national" model. Our method
for selecting parameter values for each example trajectory yields a set
of alternative models (hypotheses) about the state of a particular
caribou population; these alternative scenarios informed by prior
beliefs are used to assess the expected value of alternative sampling
strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the
simulated trajectories to confirm that the Bayesian model successfully
integrates local data with priors from the national model. For
reference, all variables and parameters are defined in supplemental
Table A1.

@@

Given the values of parameters $p_0,p_k,a,$ and $b$ used by Johnson et
al. (2020) (Table A1), recruitment rate is lowest $(0.5\dot{X}_t)$ when
$\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$ at intermediate
population sizes, and declines to $0.6\dot{X}_t$ as the population
reaches carrying capacity of $k=50000$.
The post-juvenile female population in the next year includes both
survivors and new recruits:

@@

$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example populations (Figure
\@ref(fig:exampleTrajectories)). To ensure results are comparable across
monitoring scenarios and that populations remain extant through the
monitoring period, all simulations are started with $N_0=5000$ animals
24 years prior to the start of the projection period.

@@

To assess sensitivity of results to the survival estimation method and
handle cases in which there is little or no survival data, we also
implemented an alternative parametric exponential survival model,
wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an
individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed: 
\begin{equation}
\hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).
  (\#eq:hatIEst)
\end{equation}

All regression coefficients are assumed to be Gaussian distributed, with
prior means from Johnson et al's (2020) regression model estimates
(Table \@ref(tab:priors)). We calibrated the standard deviations of the
priors so that the the 95% prior predictive intervals for survival and
recruitment from the Bayesian model match the 95% predictive intervals
from Johnson et al's regression model (Supplement C, Figure
\@ref(fig:examplePredictions)).

@@

knitr::kable(tbl,caption="Prior means and standard deviations of Gaussian distributed survival and reproduction model parameters. Standard deviations are calibrated so that the 95\\% prior predictive intervals for survival and recruitment from the Bayesian model match the 95\\% predictive intervals from Johnson et al's regression model (Supplement C).")
```

As in the simulations, we assume a simple demographic model, with a
census in late winter of surviving females and new recruits to the
post-juvenile (aged 1+) class. The number of post-juvenile females that
survive from year $t$ to the census $W_t$ is binomially distributed with
survival probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The
number of juveniles recruiting to the post-juvenile class at the census
is a binomially distributed function of the number of surviving
post-juvenile females and the adjusted recruitment rate $X_t$:
$J_{t} \sim \text{Binomial}(W_t,X_t)$. The post-juvenile female population
in the next year includes both survivors and new recruits:
$N_{t+1}=W_t+J_t$. Annual population growth rate $\lambda_t$ is set to 0
when $N_t$ is 0, and is otherwise $\lambda_t=N_{t+1}/N_t$. For more
stable estimates of population status, we report average population
growth over a period of $y=3$ years, calculated as
$\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
post-juvenile population size $N_0$ is 5000 animals. For reference, all
variables and parameters are defined in supplemental Table A1.

```{=html}
<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives biased results if there is left censoring, or months without animals at risk. To address this we drop left censored observations and only estimate survival when at least some animals remain at risk by month 12.
\*
see alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. See also Lamb et al, who uses two step KM estimation process for survival like Eacker. Stick with published examples. \*
-->
```
Models and analysis were implemented using the R language
[@r_core_team_r_2023] and a variety of *tidyverse* packages
[@wickham_welcome_2019]. We use the open source Gibbs sampling software
JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
to fit the Bayesian model. We ran 15,000 iterations of 4 chain,
discarded the first 10,000 interations, and used a thinning rate of 2.
We assessed convergence for a small random subset of example
trajectories using a threshold for the Gelman-Rubin diagnostic
[@gelman_inference_1992] of r-hat \< 1.1, and visually inspected
posterior autocorrelation and traceplots produced by the *mcmcplots* R
package [@curtis_mcmcplots_2018]. Code required to run analyses and
reproduce figures in this paper is available at
<https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>,
and in the *caribouMetrics* R package available on GitHub at
<https://github.com/LandSciTech/caribouMetrics>.

@@

In order to assess whether a monitoring program is likely sufficient to
inform decisions, it is important to select outcome measures that are
relevant to decision making [@gregory_structured_2012]. For boreal
caribou and other endangered species, a relevant outcome is the
probability of correctly predicting the true population status
(increasing or decreasing) [@dunham_strategic_2023]. To calculate this
outcome for each disturbance and monitoring scenario (Fig
\@ref(fig:figDisturbance)) we fit the Bayesian model to simulated
observations from 500 example population trajectories (e.g. Fig
\@ref(fig:exampleTrajectories)), and calculated the percentage of these
500 examples in which the Bayesian posterior mean population growth rate
and the true population growth rate over the full projection period are both greater than 0.99, or both less than or equal to 0.99.

@@

To aid in interpretation of results and to highlight the advantages of
selecting a metric that is directly relevant to the management objective
of stable or increasing populations, we also show how integrating local
monitoring data alters the distribution of differences between true
population growth rates and Bayesian posterior mean estimated growth
rates.

@@

Our analysis becomes more informative when we examine the probability of
correctly assessing population status (Figure \@ref(fig:figStatusErrors))
and the expected value of information (Figures \@ref(fig:figEVSI) and E.1).
When anthropogenic disturbance is high (right panels of Figures
\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
of population status informed only by priors from the national
demographic-disturbance relationships are accurate, and adding
additional local monitoring data does not improve accuracy.
Interestingly, when the projection period is long and anthropogenic
disturbance is high, we see a negative EVSI (lower right panel of Figure
\@ref(fig:figEVSI)), and we do not see an associated decrease in the
probability of correctly assessing population status (lower right panel
of Figure \@ref(fig:figStatusErrors)). This means that the long-term
projections informed by local monitoring data are on average less
precise but not less accurate, which can happen when there is very small
prior probability that population is viable (orange band in the top
right panel of Figure \@ref(fig:examplePredictions) falls below 1). At
lower levels of disturbance, local monitoring can reduce uncertainty,
but in some circumstances there are diminishing returns as the number of collars deployed or
the number of years of monitoring increases (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). High interannual variation makes reducing uncertainty more difficult (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).The shapes of the
relationships between monitoring effort and the probability of correct
status assessment depend on how anthropogenic disturbance is changing
over time, but there are diminishing benefits of increased monitoring
regardless of whether disturbance is increasing or decreasing
(Supplement D). Adding unknown bias increases uncertainty, but these
increases are not large for the set of parameters we investigate
(Supplement F).

```{r examplePredictions, fig.cap="Prior (a) and posterior predictions from the Bayesian model for population trajectories with low (b: $v_R=0.23,v_S=0.0435$) and high (c: $v_R=0.46,v_S=0.087$) interannual variation. In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%. The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading). Local population data (open triangles) reduces uncertainty more when interannual variation is lower (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99). When projected populations become very small, stochastic variation in projected population growth rate increases (as in the top right panel).", cache=TRUE}

@@

```{r figDiffs, fig.cap="The distribution of differences between true population growth rates and Bayesian posterior mean growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). Differences decrease as the number of years of monitoring and the number of collars increase. Differences are larger when anthropogenic disturbance is high because stochastic variation in population growth rate is higher for small populations. In these examples the composition survey cow multiplier $w$ is 6, collars are renewed each year, and interannual variation is lower ($v_R=0.23,v_S=0.0435$).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay5aSf1.png"))
```

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases. In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay5aSf1.png"))
```

```{r figEVSI, fig.cap="The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort. When disturbance is high EVSI is low, indicating that local sampling does not alter status assessments. Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but in some circumstances there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases. See Figure \\@ref(fig:figStatusErrors) for additional scenario details.", cache=TRUE}

knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay5aSf1.png"))
```

@@

Variation in anthropogenic disturbance across the Canadian boreal region
is partially correlated with variation in climate and primary
productivity [@neilson_direct_2022] because forestry occurs in higher
productivity areas, and petroleum extraction activities are concentrated
in Western Canada. Effects of disturbance on vegetation and foodweb
dynamics vary with primary productivity [@neufeld_primary_2021], and
some have expressed concern about confounding of demographic-disturbance
relationships [@wilson_causal_2021]. One option for addressing these
concerns would be to develop regional demographic-disturbance
relationships for populations that are ecologically similar to one
another (e.g. caribou inhabiting forests that are sufficiently
productive for industrial forestry in Eastern Canada)
[@eccc_scientific_2024], and use the priors from regional
demographic-disturbance relationships to inform demographic projections
within the region. However, while acknowledging and allowing for
regional variation, we also need to guard against risks that uncertainty
will be inflated and used as an excuse to delay action, a tactic that
has been used in this and other decision contexts
[@bissonnette_controverses_2020; @boan_climate_2018]. In order to be
informative and credible, regional demographic-disturbance analyses must
include a sufficient number of observations spanning a range from high
to low anthropogenic disturbance. Priors that lead to projections of
outcomes for caribou that fall outside the observed range of variation
across the country, characterized here (Figure 2) by 95% predictive
intervals from a national analysis (Johnson et al. 2020), are
implausible. In other words, if we have seen no examples of
self-sustaining caribou populations at a particularly high level of
disturbance, it is unreasonable to project or expect this outcome. These
premises constrain the plausible demographic-disturbance priors, and
clarify that no additional information is needed to predict outcomes for
caribou where anthropogenic disturbance is high (Figure 5) and no
meaningful actions are taken to support recovery.

@@

Another novel aspect of this work is a term that allows for bias
associated with misclassification of adults and missing calves in
composition surveys [@ellington_divergent_2020;
@decesare_estimating_2012]. Lacking information about the true magnitude
of this bias, we examined sensitivity to a range of assumed values, and
found that adding unknown bias did increase uncertainty, but these
increases were not large (Supplement E). A key reason for this result is
that we assumed the same range of values for the probability of
misclassifying adults and the probability of missing calves. Presuming
that surveyed groups contain more cows than bulls, and that there is an
equal probability of misclassifying cows and young bulls and vice versa,
then misclassification errors cause high estimated recruitment. Missing
calves causes low estimated recruitment, and together these errors tend
to cancel one another out. We did not investigate sensitivity to the
assumption that new recruits are 50% females, but we note that this is
another possible source of error that would most likely lead to low
recruitment estimates.

Given lack of knowledge about true error rates, the most appropriate
interpretation of our results is that better information on composition
survey error rates is required in order to assess their importance.
Fecal pellets are proving to be an interesting source of information on
sex and age ratios in surveyed groups [@mcfarlane_genetic_2018;
@jones_novel_2023]; for example, in analysis of 14 populations of boreal
and southern mountain caribou, Jones et al. 2023 found that sampled
groups contained more females than males, and that mothers tend to stay
with their daughters and sons, while fathers do not. These results are
broadly consistent with our assumption that composition survey groups
include mothers and younger animals of both sexes that can be difficult
to classify visually. Documenting animals that are difficult to classify
as “unknown” can help analysts minimize and account for classification
bias [@ellington_divergent_2020], as can comparing classification
results from multiple independent observers. The probability of missing
calves is more difficult to quantify and account for using aerial survey
methods alone, but likely increases with vegetation cover.

Alternatives to aerial composition surveys for estimating recruitment
include fecal pellet analysis [@mcfarlane_genetic_2018;
@mcfarlane_spatial_2021; @mcfarlane_identification_2022] and calf
collaring [@ellington_divergent_2020]. Each method has costs and
benefits, and the reality is that monitoring everything everywhere all
the time in great detail is neither feasible nor necessary. We hope that
a bias correction term for the simple recruitment-mortality model
[@hatter_moose_1991; @decesare_estimating_2012;
@serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011;
@johnson_science_2020] that is used for many herd viability assessments
will enable and encourage practitioners to examine the sensitivity of
their conclusions to plausible composition survey error rates, and help
clarify thinking about where, when, and why additional information is
needed.

>>>>

resultScenario = "S10"
resultSet ="cmult6ay1aSf1"
figSet = c("diffs","distScns","EVSI","power","EVPIall")

@@

```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios. Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%). These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, and not intended to reflect real landscape change. In all scenarios, the projection period begins in 2024, and the monitoring period occurs prior to the projection period.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay1aSf1.png"))

@@

To investigate the effectiveness of various monitoring strategies for
reducing uncertainty about population viability, and to show how this
varies with landscape condition, we begin with a model for simulating
example population dynamics and observations of those dynamics. Demographic parameters for the simulated trajectories are sampled from the regression models that summarize national demographic-disturbance relationships
[@johnson_science_2020] so we call this the "national" model. Together, a set of example trajectories represent alternative hypotheses about the true state of a particular caribou population; these alternative scenarios informed by prior
beliefs are used to assess the expected value of alternative sampling
strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the
simulated trajectories to confirm that the Bayesian model successfully
integrates local data with priors from the national model, and that in the absence of local data Bayesian model outcomes match expectations from the national model. For
reference, all variables and parameters are defined in supplemental
Table A1.

@@

Given the values of parameters $p_0,p_k,a,$ and $b$ used by Johnson et
al. (2020) (Table A1), recruitment rate is lowest $(0.5\dot{X}_t)$ when
$\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$ at intermediate
population sizes, and declines to $0.6\dot{X}_t$ as the population
reaches carrying capacity of $k=10000$.
The post-juvenile female population in the next year includes both
survivors and new recruits:

@@

$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example populations (Figure
\@ref(fig:exampleTrajectories)). To ensure results are comparable across
monitoring scenarios and that populations remain extant through the
monitoring period, all simulations are started with $N_0=5000$ animals
24 years prior to the start of the projection period. Realized population growth rate is $\dot{\lambda}_t=\dot{N}_t/\dot{N}_{t-1}$ Expected population growth rate (without interannual variation, demographic stochasticity, or effects of population density) is $\bar{\lambda}_t=\bar{S}_t(1+\dot{c}\bar{R}_t/2)$.

@@

To handle cases in which there is little or no survival data, we use an alternative parametric exponential survival model,
wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an
individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed: 
\begin{equation}
\hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).
  (\#eq:hatIEst)
\end{equation}

All regression coefficients are assumed to be Gaussian distributed, with
prior means from Johnson et al's (2020) regression model estimates
(Table \@ref(tab:priors)). We calibrated the standard deviations of the
priors so that the the 95% prior predictive intervals for survival and
recruitment from the Bayesian model match the 95% prediction intervals
from Johnson et al's regression model (Supplement C, Figure
\@ref(fig:examplePredictions)).

@@

knitr::kable(tbl,caption="Prior means and standard deviations of Gaussian distributed survival and reproduction model parameters. Standard deviations are calibrated so that the 95\\% prior predictive intervals for survival and recruitment from the Bayesian model match the 95\\% prediction intervals from Johnson et al's regression model (Supplement C).")
```

As in the simulations, we assume a simple demographic model, with a
census in late winter of surviving females and new recruits to the
post-juvenile (aged 1+) class. Expected population growth rate (without interannual variation or demographic stochasticity) is $\tilde{\lambda}_t=\tilde{S}_t(1+c\tilde{R}_t/2)$. For reference, all variables and parameters are defined in supplemental Table A1.

Models and analysis were implemented using the R language
[@r_core_team_r_2023] and a variety of *tidyverse* packages
[@wickham_welcome_2019]. We use the open source Gibbs sampling software
JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
to fit the Bayesian model. We ran 15,000 iterations of 4 chain,
discarded the first 10,000 interations, and used a thinning rate of 2.
We assessed convergence for a small random subset of example
trajectories using a threshold for the Gelman-Rubin diagnostic
[@gelman_inference_1992] of r-hat \< 1.1, and visually inspected
posterior autocorrelation and traceplots produced by the *mcmcplots* R
package [@curtis_mcmcplots_2018]. Code required to run analyses and
reproduce figures in this paper is available at
<https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>,
and in the *caribouMetrics* R package available on GitHub at
<https://github.com/LandSciTech/caribouMetrics>.

@@

In order to assess whether a monitoring program is likely sufficient to
inform decisions, it is important to select outcome measures that are
relevant to decision making [@gregory_structured_2012]. For boreal
caribou and other endangered species, a relevant outcome is the
probability of correctly predicting the true population status
(increasing or decreasing) [@dunham_strategic_2023]. To calculate this
outcome for each disturbance and monitoring scenario (Fig
\@ref(fig:figDisturbance)) we fit the Bayesian model to simulated
observations from 500 example population trajectories (e.g. Fig
\@ref(fig:exampleTrajectories)), and calculated the percentage of these
500 examples in which the Bayesian posterior expected population growth rate
and the true expected population growth rate are both greater than 0.99, or both less than or equal to 0.99.

@@

To aid in interpretation of results and to highlight the advantages of
selecting a metric that is directly relevant to the management objective
of stable or increasing populations, we also show how integrating local
monitoring data alters the distribution of differences between true
expected population growth rates and Bayesian posterior expected growth
rates.

@@

Our analysis becomes more informative when we examine the probability of
correctly assessing population status (Figure \@ref(fig:figStatusErrors))
and the expected value of information (Figures \@ref(fig:figEVSI) and E.1).
When anthropogenic disturbance is high (right panels of Figures
\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
of population status informed only by priors from the national
demographic-disturbance relationships are accurate, and adding
additional local monitoring data does not improve accuracy.
Interestingly, when the projection period is long and anthropogenic
disturbance is high, we see a negative EVSI (lower right panel of Figure
\@ref(fig:figEVSI)), and we do not see an associated decrease in the
probability of correctly assessing population status (lower right panel
of Figure \@ref(fig:figStatusErrors)). This means that the long-term
projections informed by local monitoring data are less
precise but not less accurate, which can happen if the prior viability predictions are more certain than the posterior predictions (e.g 
left side of right panels in Figure \@ref(fig:examplePredictions)). At
lower levels of disturbance, local monitoring can reduce uncertainty,
there are diminishing returns as the number of collars deployed or
the number of years of monitoring increases (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). Reducing uncertainty is more difficult when interannual variation is high (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).The shapes of the
relationships between monitoring effort and the probability of correct
status assessment depend on how anthropogenic disturbance is changing
over time, but there are diminishing benefits of increased monitoring
regardless of whether disturbance is increasing or decreasing
(Supplement D). Adding unknown bias increases uncertainty, but these
increases are not large for the set of parameters we investigate
(Supplement F).

```{r examplePredictions, fig.cap="Prior (a) and posterior predictions from the Bayesian model for population trajectories with low (b: $v_R=0.23,v_S=0.0435$) and high (c: $v_R=0.46,v_S=0.087$) interannual variation. In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%. The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading). Local population data (open triangles) reduces uncertainty about demographic rates, but local data is not required to predict viability (population growth <= 0.99) on highly disturbed landscapes.", cache=TRUE}

@@

```{r figDiffs, fig.cap="The distribution of differences between true expected population growth rates and Bayesian posterior expected growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). Differences decrease as the number of years of monitoring and the number of collars increase. In these examples the composition survey cow multiplier $w$ is 6, collars are renewed each year, and interannual variation is lower ($v_R=0.23,v_S=0.0435$).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay1aSf1.png"))
```

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as the number of years or collars increases. In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay1aSf1.png"))
```

```{r figEVSI, fig.cap="The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort. When disturbance is high EVSI is low, indicating that local sampling does not change predicted status.Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring (number of years) as the number of years increases. Negative EVSI values indicate that the posterior population viability prediction is less certain than the prior prediction. See Figure \\@ref(fig:figStatusErrors) for additional scenario details.", cache=TRUE}

knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay1aSf1.png"))
```

@@

Variation in anthropogenic disturbance across the Canadian boreal region
is partially correlated with variation in climate and primary
productivity [@neilson_direct_2022] because forestry occurs in higher
productivity areas, and petroleum extraction activities are concentrated
in Western Canada. Effects of disturbance on vegetation and foodweb
dynamics vary with primary productivity [@neufeld_primary_2021], and
some have expressed concern about confounding of demographic-disturbance
relationships [@wilson_causal_2021]. One option for addressing these
concerns would be to develop regional demographic-disturbance
relationships for populations that are ecologically similar to one
another (e.g. caribou inhabiting forests that are sufficiently
productive for industrial forestry in Eastern Canada)
[@eccc_scientific_2024], and use the priors from regional
demographic-disturbance relationships to inform demographic projections
within the region. However, while acknowledging and allowing for
regional variation, we also need to guard against risks that uncertainty
will be inflated and used as an excuse to delay action, a tactic that
has been used in this and other decision contexts
[@bissonnette_controverses_2020; @boan_climate_2018]. In order to be
informative and credible, regional demographic-disturbance analyses must
include a sufficient number of observations spanning a range from high
to low anthropogenic disturbance. Priors that lead to projections of
outcomes for caribou that fall outside the observed range of variation
across the country, characterized here (Figure 2) by 95% prediction
intervals from a national analysis (Johnson et al. 2020), are
implausible. In other words, if we have seen no examples of
self-sustaining caribou populations at a particularly high level of
disturbance, it is unreasonable to project or expect this outcome. These
premises constrain the plausible demographic-disturbance priors, and
clarify that no additional information is needed to predict outcomes for
caribou where anthropogenic disturbance is high (Figure 5) and no
meaningful actions are taken to support recovery.

@@

Another novel aspect of this work is a term that allows for bias
associated with misclassification of adults and missing calves in
composition surveys [@ellington_divergent_2020;
@decesare_estimating_2012]. Lacking information about the true magnitude
of this bias, we examined sensitivity to a range of assumed values, and
found that adding unknown bias did increase uncertainty, but these
increases were not large (Supplement E). A key reason for this result is
that we assumed the same range of values for the probability of
misclassifying adults and the probability of missing calves. Presuming
that surveyed groups contain more cows than bulls, and that there is an
equal probability of misclassifying cows and young bulls and vice versa,
then misclassification errors cause high estimated recruitment. Missing
calves causes low estimated recruitment, and together these errors tend
to cancel one another out.

Given lack of knowledge about true error rates, the most appropriate
interpretation of our results is that better information on composition
survey error rates is required in order to assess their importance.
Fecal pellets are proving to be an interesting source of information on
sex and age ratios in surveyed groups [@mcfarlane_genetic_2018;
@jones_novel_2023]; for example, in analysis of 14 populations of boreal
and southern mountain caribou, Jones et al. 2023 found that sampled
groups contained more females than males, and that mothers tend to stay
with their daughters and sons, while fathers do not. These results are
broadly consistent with our assumption that composition survey groups
include mothers and younger animals of both sexes that can be difficult
to classify visually. Documenting animals that are difficult to classify
as “unknown” can help analysts account for classification uncertainties [@ellington_divergent_2020;@dalgarno_bboutools_2024], as can comparing classification
results from multiple independent observers. The probability of missing
calves is more difficult to quantify and account for using aerial survey
methods alone, but likely increases with vegetation cover.

Alternatives to aerial composition surveys for estimating recruitment
include fecal pellet analysis [@mcfarlane_genetic_2018;
@mcfarlane_spatial_2021; @mcfarlane_identification_2022] and calf
collaring [@ellington_divergent_2020]. Each method has costs and
benefits, and the reality is that monitoring everything everywhere all
the time in great detail is neither feasible nor necessary. We hope that
a bias correction term for the simple recruitment-
mortality model that is used for many herd viability assessments [@hatter_moose_1991; @decesare_estimating_2012;
@serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011;
@johnson_science_2020; @dalgarno_bboutools_2024] will enable and encourage practitioners to examine the sensitivity of their conclusions to plausible composition survey error rates, and help clarify thinking about where, when, and why additional information is
needed.
